Ubiquitous Digital Technologies and Spatial Structure; an update

### 2020-12-02 paper/2_resubmission/tranos_ioannides.Rmd
minor changes

\color{blue}
Kolko [@kolko_death_2000] uses internet diffusion data and identifies a clear complementary link between internet usage and city size.
Interestingly, he identifies higher internet domain densities in remote cities which indicated a substitution effect of the internet for longer-distance non-electronic communications. His results are consistent for different measures of internet diffusion (internet domain density and internet take-up).
Sinai and Waldfogel [@todd_sinai_geography_2004] approach the same question from the consumers’ point of view and find support for complementarity between internet and cities.
They study the link between market size and locally targeted online content and find that more online local content in  larger markets.
However, they also obtain evidence for a substitution effect: they find that holding constant local online content constant, market size has a negative effect on individual connectivity.
Forman et al. [@chris_forman_how_2005] examined whether commercial internet adoption is higher in cities than in rural areas. While the former would indicate a complementarity between internet adoption and cities, the latter would reflect a complementary relation, according to which the internet is used as a means to offset costs and lack of opportunities related to peripheriality.
Their results indicate that despite internet adoption by firms with more than 100 employees being greater in smaller urban agglomerations, the adoption of more sophisticated internet-based applications is positively related with city size in $2000$.
Sohn, Kim, and Hewings [@jungyul_sohn_information_2003] compare how information technologies are related to urban spatial structure for Chicago and Seoul.
Although they find a clear complementary link for Chicago, this was not the case for Seoul, where information technologies contribute to a more dispersed spatial pattern.
Focusing on the municipalities in the province of Barcelona, Pons-Novell and Viladecans-Marsal [@pons-novell_cities_2006] find a complementary link between individual internet take-up and off-line commercial offerings, but their results cannot safely reject a substitution effect.
Craig, Hoang, and Kohlhase [@craig_does_2016] focus on internet take-up rates across US states during the period $2000-2011$.
Their analysis provides suggestive evidence of a complementary role that internet connectivity performs on urban living.
Ioannides et al. [@ioannides2008effect] use country-level data on city size distributions to examine the impact of fixed line telephony on urban structure $1980-2000$.
Using a panel dataset of spatial dispersion measures, they find robust evidence that an increase in the number of telephone lines per capita encourages the spatial dispersion of population in that they lead to a more concentrated distribution of city sizes.
In addition, by the end of their study period the internet has come into use, but their results with internet usage is more speculative but do show that it goes in the same direction.

@@

\color{blue}
The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue, as it is clear that not all economic activities might continue in an
unencumbered manner; see Dingel and Neiman [@dingel2020many].
\color{black}

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
\color{blue}
Pareto exponent coefficient
\color{black}
for a broad sample of countries over time.
\color{blue}
The Pareto coefficient
\color{black}
is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations^[
\color{blue}
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.].
\color{black}
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
\color{blue}
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engage with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
\color{black}
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

\color{blue}
It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
does not make up for physical distance; individuals nearer one another
communicate more.  
\color{black}

@@

The LHS variable we use here is the outcome of equation \ref{eq5}.
\color{blue}
We use this variable to estimate \ref{eq4} with OLS.
Additionally, we employ a normalized form of the dependent variable, so as that
variable is bounded between $0$ and $1$ and with a mean of  $0.5$^[$$Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.$$].
This transformation enables us to estimate \ref{eq4}
with a quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a difference between
two count variables, which may also assume negative values.
\color{blue}
The results of the different estimations remain qualitatively the same regardless of the form of the LHS variable we use.
Table \ref{ols.usa} reports the results of the estimation of equation \ref{eq4} using OLS.
The sign and significance of the main variable of interest -- percentage of population with computer and broadband connection -- verifies our global level results.
That is an increase in internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that are OLS and GLM.
In addition, Table \ref{ols.usa} also reports the estimates of interaction effects between the share of population with broadband connection with population and population density.
Although the interaction terms are not significant, the main effect remains qualitatively unchanged.

@@

Although not directly comparable, the above estimations are in accordance with our global model. Moreover, the potential presence of endogeneity might be a problem here as well.
Therefore, Table \ref{2sls.usa} reports 2SLS estimations.
In addition to the strategy employed for the cross-country analysis -- that is the inclusion of one instrument for which we have strong reasons to believe is uncorrelated with the error term -- data availability allows to employ a second IV in order to estimate the Sargan over-identification restrictions test (Column 2).
The main instrument we propose here is Bachelor’s degrees per inhabitant in $2005$^[
\color{blue}
Due to differences in scale, time and the endogenous variables, different IVs
have been unutilized for the different sections of the paper.
\color{black}].
Even if the quality of human capital can affect the population growth of a city $10$ years later, our LHS variable adopts a systemic understanding of the US urban system as it measures the relative position of a city within the overall urban system instead of its population growth.
In other words, a city might experience population growth between two periods, but if other cities have also experienced population growth, this might not affect its relative position within the urban system.
Hence, we do not expect that this IV affects the LHS variable.
<!-- In addition, the correlation coefficient between these two variables is still very low ($r=0.24$). -->
Moreover, when we add the second IV, which in this case is the commute time in minutes also in $2005$, we fail to reject the null hypothesis of the Sargan test, something which also adds on the validity of our strategy.
Furthermore, our results do not suffer from weak identification according to the relevant tests in Table \ref{2sls.usa}.

@@

In regards to IV strategy, we use here the *number of universities* as the main instrument.
Universities in England and Wales have been established long time ago and therefore we do not expect them to affect the change in the relative position of built-up areas in $2011$.
\color{blue}
The presence of universities and large student populations tend to be correlated
with the quality of internet infrastructure provision [@GRUBESIC201950].
\color{black}
<!-- The correlation coefficient with the LHS variables is smaller than $0.05$. -->
The addition of the second instrument (absolute number of broadband tests in 2011) in column $2$ results to a Sargan test, the null hypothesis of which cannot be rejected.
Moreover, the estimations reported in Table 11 do not suffer from weak instruments as the relevant test is above the rule of thumb value of $10$.
\color{blue}
The coefficients derived from OLS (Table \ref{ols.uk}) and 2SLS (Table \ref{2sls.uk}) are always positive and significant, something which advocates towards the effect of broadband speed in enhancing agglomeration forces.
Interestingly, the magnitude of the 2SLS coefficient (column 1 in Table \ref{2sls.uk})
is almost 8 times higher than the OLS coefficient (column 1 in Table \ref{ols.uk}).
We attribute the OLS underestimation, which is more or less consistent in all
our estimations, to the difficulty in directly observing active engagement with
ICT. Our endogenous variable captures the quality of the internet infrastructure
as the experienced internet speed.
Similarly to the endogenous variables of the global and the US model (internet
and telephony take-up), what these variables do not capture is what individuals
actually do with ICT [see @tranos2020digital for such an application]. For example, if we assume a bi-modal internet usage for
either entertainment or work purposes, one would expect  that the latter would
have a larger effect on spatial structure than the former. We believe that our
endogenous variables capture mostly the former. Therefore, the IV estimations,
especially for UK case, reveal the local average treatment effect
[@angrist1995identification] by capturing the effect of these users/households
who actively engage with ICT (e.g. work on ICT-related sectors) and, therefore, have a higher effect on spatial structure.
A back of the envelope calculation indicates that, everything else being equal,
$10$ per cent increase in download speed for a BUA,
would have increased its position by $4$ places.
\color{black}

>>>>

\color{blue}
Kolko [@kolko_death_2000] uses internet diffusion data and identifies a clear complementary link between internet usage and city size.
Interestingly, he identifies higher internet domain densities in remote cities which indicates a substitution effect of the internet for longer-distance non-electronic communications. His results are consistent for different measures of internet diffusion (internet domain density and internet take-up).
Sinai and Waldfogel [@todd_sinai_geography_2004] approach the same question from the consumers’ point of view and find support for complementarity between internet and cities.
They study the link between market size and locally targeted online content and find more online local content in  larger markets.
However, they also obtain evidence for a substitution effect: holding local online content constant, market size has a negative effect on individual connectivity.
Forman et al. [@chris_forman_how_2005] examines whether commercial internet adoption is higher in cities than in rural areas. While the former would indicate a complementarity between internet adoption and cities, the latter would reflect a substitutability relation, according to which the internet is used as a means to offset costs and lack of opportunities related to peripheriality.
Their results indicate that despite internet adoption by firms with more than 100 employees being greater in smaller urban agglomerations, the adoption of more sophisticated internet-based applications is positively related with city size in $2000$.
Sohn, Kim, and Hewings [@jungyul_sohn_information_2003] compare how information technologies are related to urban spatial structure for Chicago and Seoul.
Although they find a clear complementary link for Chicago, this is not the case for Seoul, where information technologies contribute to a more dispersed spatial pattern.
Focusing on the municipalities in the province of Barcelona, Pons-Novell and Viladecans-Marsal [@pons-novell_cities_2006] find a complementary link between individual internet take-up and off-line commercial offerings, but their results cannot safely reject a substitution effect.
Craig, Hoang, and Kohlhase [@craig_does_2016] focus on internet take-up rates across US states during the period $2000-2011$.
Their analysis provides suggestive evidence of a complementary role that internet connectivity performs on urban living.
Ioannides et al. [@ioannides2008effect] use country-level data on city size distributions to examine the impact of fixed line telephony on urban structure $1980-2000$.
Using a panel dataset of spatial dispersion measures, they find robust evidence that an increase in the number of telephone lines per capita encourages the spatial dispersion of population in that they lead to a more concentrated distribution of city sizes.
In addition, by the end of their study period the internet has come into use, but their results with internet usage is more speculative but do show that it goes in the same direction.

@@

\color{blue}
The present paper does not address the issue that adoption of ICT may have
different effects on urban dispersion on a national scale, that is how far are
major urban areas, that is, cities, from one another, versus dispersion within major
urban areas, that is urban sprawl. Such an inquiry would require detailed data
about patterns of how urban sprawl may hamper or promote economic interaction,
given advancements in information technology.
We think that data from the Covid-19 pandemic would likely throw light at this
issue in the future, as it is clear that not all economic activities might
continue in an unencumbered manner; see Dingel and Neiman [@dingel2020many].
\color{black}

@@

The multi-country identification strategy is a two-step approach [@ioannides2008effect].
The first step of our methodology is to estimate the
\color{blue}
Pareto exponent
\color{black}
for a broad sample of countries over time.
\color{blue}
The Pareto exponent
\color{black}
is one of the most widely used measures of spatial dispersion with numerous applications in urban economics and economic geography ([see for example @black_urban_2003; @frenken_theoretical_2007; @giesen_zipfs_2011; @rauch_cities_2013; @ioannides_zipfs_2003; @ioannides_walled_2017; @nitsch_zipf_2005).
This an appropriate measure of dispersion because of the extreme heterogeneity of the city size distribution and the very good fits normally obtained with such estimations^[
\color{blue}
We also replicate our analysis using other measures of dispersion in Appendix S1. Our results remain consistent.].
\color{black}
Skipping details pertaining of the suitability of the approach, which may be found in Ioannides et al. [@ioannides2008effect] and Tranos and Ioannides [@tranos2020ict], we proceed with the estimation of the logarithmic form of:

@@

The estimations presented in Table \ref{2sls.global} indicate that increases in internet and broadband internet usage have resulted to a decrease of the spatial dispersion of population for the time period and for the panel of countries included in our data.
In other words, increases in internet and digital communications resulted to national urban systems which are less uniform in terms of city sizes and are characterized by higher population concentrations in larger cities.
The exact opposite appears to happen for fixed telephony, which our results indicate that has led to an increase of spatial dispersion of population.
\color{blue}
We expect that the IV estimates capture the local average treatment effect
[@angrist1995identification] of these labour markets, which are characterised by
more active engagement with the internet, and, therefore, experience a larger effect
on the Pareto exponents.
\color{black}
Importantly, our identification strategy enables us to address potential reverse causality issues and treat the results of Table \ref{2sls.global} as causal. Hence, the main finding of our multi-country, global analysis based on data for urban agglomerations of at least 300,000 inhabitants is that internet and broadband internet usage appear to act in favor of agglomeration economies and result to urban systems with more dominant cities on the top of the urban hierarchy.
On the contrary, fixed telephony usage acted in favour of centrifugal forces.
Our identification strategy does not enable us to make similar interpretations for mobile telephony usage.

@@

\color{blue}
It is indeed puzzling that the results on Table \ref{2sls.global} that the
effects on urban concentration for internet users and broadband users, on the
one hand, and mobile and fixed telephony users are in opposite directions.
We wish to broach this subject in the context of some related literature,
namely [@calabrese2011interplay] and [@buchel2020cities].
The latter supports the notion that ICT favors agglomeration in the sense that
in spite of their convenience modern technologies do no make up for the need of
proximity in  social interactions.
The former supports the notion people who communicate more frequently are likely
to be near one another.
ICT, as viewed through such granular studies of individuals' communications
do not make up for physical distance; individuals nearer one another
communicate more.  
\color{black}

@@

The LHS variable we use here is the outcome of equation \ref{eq5}.
\color{blue}
We use this variable to estimate \ref{eq4} with OLS.
Additionally, we employ a normalized form of the dependent variable, so as that
variable is bounded between $0$ and $1$ and with a mean of  $0.5$^[$$Normalized\:difference\:in\:ranks_{i} = \frac{difference\:in\:ranks_{i} + N_{cities}}{2xN_{cities}}.$$].
This transformation enables us to estimate \ref{eq4}
with a quasibinomial GLM estimator given that the original form of our LHS
variable does not vary continuously, but is instead defined as a difference between
two count variables, which may also assume negative values.
\color{blue}
The results of the different estimations remain qualitatively the same regardless of the form of the LHS variable we use.
Table \ref{ols.usa} reports the results of the estimation of equation \ref{eq4} using OLS.
The sign and significance of the main variable of interest -- percentage of population with computer and broadband connection -- verifies our global level results.
That is, an increase in internet usage improves the position of a city in the US urban system.
The results are consistent among the different estimators that is OLS and GLM.
In addition, Table \ref{ols.usa} also reports the estimates of interaction effects between the share of population with broadband connection with population and population density.
Although the interaction terms are not significant, the main effect remains qualitatively unchanged.

@@

Although not directly comparable, the above estimations are in accordance with our global model. Moreover, the potential presence of endogeneity might be a problem here as well.
Therefore, Table \ref{2sls.usa} reports 2SLS estimations.
In addition to the strategy employed for the cross-country analysis -- that is the inclusion of one instrument for which we have strong reasons to believe is uncorrelated with the error term -- data availability allows us to employ a second IV in order to estimate the Sargan over-identification restrictions test (Column 2).
The main instrument we propose here is Bachelor’s degrees per inhabitant in $2005$^[
\color{blue}
Due to differences in scale, time and the endogenous variables, different IVs
have been unutilized for the different sections of the paper.
\color{black}].
Even if the quality of human capital affect the population growth of a city $10$ years later, our LHS variable adopts a systemic understanding of the US urban system as it measures the relative position of a city within the overall urban system instead of its population growth.
In other words, a city might experience population growth between two periods, but if other cities have also experienced population growth, this might not affect its relative position within the urban system.
Hence, we do not expect that this IV affects the LHS variable.
<!-- In addition, the correlation coefficient between these two variables is still very low ($r=0.24$). -->
Moreover, when we add the second IV, which in this case is the commute time in minutes also in $2005$, we fail to reject the null hypothesis of the Sargan test, something which also adds the validity of our strategy.
Furthermore, our results do not suffer from weak identification according to the relevant tests in Table \ref{2sls.usa}.

@@

In regards to IV strategy, we use here the *number of universities* as the main instrument.
Universities in England and Wales have been established long time ago and therefore we do not expect them to affect the change in the relative position of built-up areas in $2011$.
\color{blue}
The presence of universities and large student populations tend to be correlated
with the quality of internet infrastructure provision [@GRUBESIC201950].
\color{black}
<!-- The correlation coefficient with the LHS variables is smaller than $0.05$. -->
The addition of the second instrument (absolute number of broadband tests in 2011) in column $2$ results to a Sargan test, the null hypothesis of which cannot be rejected.
Moreover, the estimations reported in Table 11 do not suffer from weak instruments as the relevant test is above the rule of thumb value of $10$.
\color{blue}
The coefficients derived from OLS (Table \ref{ols.uk}) and 2SLS (Table \ref{2sls.uk}) are always positive and significant, something which advocates in favour of the effect of broadband speed in enhancing agglomeration forces.
Interestingly, the magnitude of the 2SLS coefficient (column 1 in Table \ref{2sls.uk})
is almost 8 times higher than the OLS coefficient (column 1 in Table \ref{ols.uk}).
We attribute the OLS underestimation, which is more or less consistent in all
our estimations, to the difficulty in directly observing active engagement with
ICT. Our endogenous variable captures the quality of the internet infrastructure
as the experienced internet speed.
Similarly to the endogenous variables of the global and the US model (internet
and telephony take-up), what these variables do not capture is what individuals
actually do with ICT [see @tranos2020digital for such an application]. For example, if we assume a bi-modal internet usage for
either entertainment or work purposes, one would expect  that the latter would
have a larger effect on spatial structure than the former. We believe that our
endogenous variables capture mostly the former. Therefore, the IV estimations,
especially for UK case, reveal the local average treatment effect
[@angrist1995identification] by capturing the effect of these users/households
who actively engage with ICT (e.g. work on ICT-related sectors) and, therefore, have a higher effect on spatial structure.
A back of the envelope calculation indicates that, everything else being equal,
$10$ per cent increase in download speed for a BUA,
would have increased its position by $4$ places.
\color{black}
